Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-20 arxiv-version/main.tex
Small fixes

We design new algorithms for the problem of \textsc{Online Multiclass
Classification with Bandit Feedback}. In this problem, in each round, the
algorithm receives a feature vector and algorithm has to predict one of $K$
possible classes. Imediately after the algorithm makes its prediction, it
receives feedback whether the prediction was correct or not. However, in
contrast with the standard, so called full information setting, the algorithm
does \emph{not} receive the correct label. The feedback is only binary, correct
prediction or incorrect prediction. To allow for meaningful randomized
strategies, we assume that the correct label in a particular round is chosen
prior to algorithm's prediction in that round.\footnote{The reader can assume
throughout the paper a simpler model of the so called \emph{oblivious adversary}
where the whole sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots,
(x_T, y_T)$ is chosen before the first round of the game.} The protocol of the
problem is stated below.

@@

The bandit feedback setting, however, is much more challenging. To understand
the differences, we dissect the notion of linear separability in the multiclass
setting. Beside the standard notion of linear separability, we define also
\emph{strong linear separability}. For binary classification, these two
definitions are identical. However, they differ for multiclass classification.
For multiclass classification with $K$ classes the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class can be separated from the remaining examples by a
\emph{single} hyperplane. We state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

@@

\begin{definition}
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, and let $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ be labeled examples in $V \times \{1,2,\dots,K\}$.

@@

Similarly, if $K=2$ and labeled examples are linearly separable with a margin
$\gamma$ then they are strongly linearly separable with margin $\gamma/2$.
Indeed, if $w_1, w_2$ satisfy \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} then $w_1' = \frac{w_1 - w_2}{2}$,
$w_2' = \frac{w_2 - w_1}{2}$ and $\gamma'=\gamma/2$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} since $\norm{w_i'}^2 \le
(\frac{1}{2} \norm{w_1} + \frac{1}{2} \norm{w_2})^2 \le \frac{1}{2}\norm{w_1}^2 +
\frac{1}{2}\norm{w_2}^2$ for $i=1,2$.

@@

While the examples are linearly separable with a positive margin $\gamma$, they
are \emph{not} strongly linearly separable for any positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

@@

In this section, we present an algorithm for \textsc{Online Multiclass
Classification with Bandit Feedback} and prove an upper bound on the number of
mistakes under the assumption that the examples are strongly linearly separable
with a margin. The algorithm is stated as
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} and
the mistake upper bound is stated as
\autoref{theorem:stronglys-separable-example-mistake-upper-bound}.

@@

\begin{theorem}
\label{theorem:stronglys-separable-example-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, $\gamma$ a positive real number, $R$ be a non-negative real number If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples in
$V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then the
expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

Let $C$ be the number of times line 16 gets executed. Let $U$ be the number of
times line 16 or line 21 get executed. In other words, $U$ is the number of
times the weight vectors are updated. Clearly, $U = A + C$.

The key observation is that $\Exp[B] = (K-1) \Exp[C]$ since if $S_t =
\emptyset$, there is $1/K$ probability that the algorithm guesses the correct
label correctly and $(K-1)/K$ probability that it guesses is incorrectly.
Putting all the information together, we get that
\begin{align*}
\Exp[M]
& = \Exp[A] + \Exp[B] \\
& = \Exp[A] + (K-1) \Exp[C]  \\
& \le (K-1) \Exp[A + C] \\
& = (K-1) \Exp[U]  \; .
\end{align*}

To finish the proof, we need to upper bound the number of updates, $U$. We claim
that $U \le \lfloor (R/\gamma)^2 \rfloor$. The proof of this upper bound is
similar to the proof of the mistake bound for \textsc{Multiclass Perceptron}
algorithm. Let $w_1^*, w_2^*, \dots, w_K^* \in V$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}.
The $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$
changes only if there is an update in round $t$.
We investigate how $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ and
$\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$ change. If there is an update in round $t$,
we have
\begin{align*}
\sum_{i=1}^K \norm{w_i^{(t+1)}}^2
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t}^2 \\
& = \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 + \underbrace{(-1)^{z_t} 2 \ip{w_{\widehat y_t}^{(t)}}{x_t}}_{\le 0} \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + R^2 \; .
\end{align*}
Therefore, after $U$ updates,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
$$
Similarly, if there is an update in round $t$, we have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + (-1)^{z_t} \ip{w_{\widehat y_t}^*}{x_t} \\
& \ge \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma
\end{align*}
where the last inequality follows from \eqref{equation:strong-linear-separability-2}
and \eqref{equation:strong-linear-separability-3} and case analysis whether $z_t = 0$ or $z_t = 1$.
Thus, after $U$ updates,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma U \; .
$$
Applying Cauchy-Schwartz's inequality twice, we get that
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
\le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}}
\le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2}
\le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
$$
Combining the upper and lower bounds on $\sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2}$ inequalities, we get
$$
(\gamma U)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
$$
We conclude that $U \le (R/\gamma)^2$. Since $U$ is an integer, $U \le \lfloor (R/\gamma)^2 \rfloor$.
\end{proof}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Full-Information Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

>>>>

We design new algorithms for the problem of \textsc{Online Multiclass
Classification with Bandit Feedback}. In this problem, in each round, the
algorithm receives a feature vector and algorithm has to predict one of $K$
possible classes. Imediately after the algorithm makes its prediction, it
receives feedback whether the prediction was correct or not. However, in
contrast with the standard, so called full information setting, the algorithm
does \emph{not} receive the correct label. The feedback is only binary. To allow
for meaningful randomized strategies, we assume that the correct label in a
particular round is chosen prior to algorithm's prediction in that
round.\footnote{The reader can assume throughout the paper a simpler model of
the so called \emph{oblivious adversary} where the whole sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is chosen before the first
round of the game.} The protocol of the problem is stated below.

@@

The bandit feedback setting, however, is much more challenging. To understand
the differences, we dissect the notion of linear separability in the multiclass
setting. Beside the standard notion of linear separability, we define also
\emph{strong linear separability}. For binary classification, these two
definitions are identical. However, they differ for multiclass classification.
For multiclass classification with $K$ classes the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class is separated from the remaining examples by a
\emph{single} hyperplane. We state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

@@

\begin{definition}[Linear separability]
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, and let $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ be labeled examples in $V \times \{1,2,\dots,K\}$.

@@

Similarly, if $K=2$ and labeled examples are linearly separable with a margin
$\gamma$ then they are strongly linearly separable with margin $\gamma/2$.
Indeed, if $w_1, w_2$ satisfy \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} then $w_1' = \frac{w_1 - w_2}{2}$, $w_2' =
\frac{w_2 - w_1}{2}$ and $\gamma'=\gamma/2$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3}. The last two conditions are easy
to see. The first condition follows from $\norm{w_i'}^2 \le (\frac{1}{2}
\norm{w_1} + \frac{1}{2} \norm{w_2})^2 \le \frac{1}{2}\norm{w_1}^2 +
\frac{1}{2}\norm{w_2}^2 \le \frac{1}{2}$ for $i=1,2$.

@@

While the examples are linearly separable with a positive margin $\gamma$, they
are \emph{not} strongly linearly separable with any positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

@@

In this section, we present an algorithm for \textsc{Online Multiclass
Classification with Bandit Feedback} and prove an upper bound on the number of
mistakes under the assumption that the examples are strongly linearly separable
with a margin. The algorithm is stated as
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} and
the mistake upper bound is stated as
\autoref{theorem:strongly-separable-example-mistake-upper-bound}.

@@

\begin{theorem}[Mistake upper bound]
\label{theorem:strongly-separable-example-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, $\gamma$ a positive real number, $R$ be a non-negative real number If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples in
$V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then the
expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

Let $C$ be the number of times line 16 gets executed. Let $U$ be the number of
times line 16 and line 21 get executed. In other words, $U$ is the number of
times the $K$-tuple of vectors $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$ gets
updated. Clearly, $U = A + C$.

The key observation is that $\Exp[B] = (K-1) \Exp[C]$ since if $S_t =
\emptyset$, there is $1/K$ probability that the algorithm guesses the correct
label and with probability $(K-1)/K$ algorithm's guess is incorrect.
Putting all the information together, we get that
\begin{align*}
\Exp[M]
& = \Exp[A] + \Exp[B] \\
& = \Exp[A] + (K-1) \Exp[C]  \\
& \le (K-1) \Exp[A + C] \\
& = (K-1) \Exp[U]  \; .
\end{align*}

To finish the proof, we need to upper bound the number of updates, $U$. We claim
that $U \le \lfloor (R/\gamma)^2 \rfloor$. The proof of this upper bound is
similar to the proof of the mistake bound for \textsc{Multiclass Perceptron}
algorithm. Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors that satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}.
The $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$
changes only if there is an update in round $t$.
We investigate how $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ and
$\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$ change. If there is an update in round $t$,
we have
\begin{align*}
\sum_{i=1}^K \norm{w_i^{(t+1)}}^2
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t}^2 \\
& = \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 + \underbrace{(-1)^{z_t} 2 \ip{w_{\widehat y_t}^{(t)}}{x_t}}_{\le 0} \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + R^2 \; .
\end{align*}
Therefore, after $U$ updates,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
$$
Similarly, if there is an update in round $t$, we have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + (-1)^{z_t} \ip{w_{\widehat y_t}^*}{x_t} \\
& \ge \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma
\end{align*}
where the last inequality follows from \eqref{equation:strong-linear-separability-2}
and \eqref{equation:strong-linear-separability-3} and case analysis whether $z_t = 0$ or $z_t = 1$.
Thus, after $U$ updates,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma U \; .
$$
Applying Cauchy-Schwartz's inequality twice and assumption
\eqref{equation:strong-linear-separability-1}, we get that
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
\le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}}
\le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2}
\le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
$$
Combining the upper and lower bounds on $\sum_{i=1}^K \norm{w_i^{(T+1)}}^2$ inequalities, we get
$$
(\gamma U)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
$$
We conclude that $U \le (R/\gamma)^2$. Since $U$ is an integer, $U \le \lfloor (R/\gamma)^2 \rfloor$.
\end{proof}

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-example-mistake-lower-bound} Let $K$ be a
positive integer, let $\gamma$ be a positive real number and let $R$ be a
non-negative real number. For any algorithm for the \textsc{Online Multiclass
Classification with Bandit Feedback} there exists an inner product space $(V,
\ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ examples in $V \times
\{1,2,\dots,K\}$ that are strongly linearly separable with margin $\gamma$, the
norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the expected
number mistakes the algorithm is at least $K \lfloor (R/\gamma)^2 \rfloor/ ??$.
\end{theorem}

\begin{proof}
TODO
\end{proof}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Online Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}
